---
title: "Correlation"
output: html_notebook
---

Clear environment and load libraries


```{r}
rm(list = ls())

library('dplyr')
library('ggplot2')
```


Load and manipulate data

```{r}
surveys <- read.csv("data/combined.csv")
surveys_complete <- surveys %>%
  filter(species_id != "",         # remove missing species_id
         !is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         sex != "")                # remove missing sex

# Extract the most common species_id
species_counts <- surveys_complete %>%
  group_by(species_id) %>%
  tally %>%
  filter(n >= 50)

# Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)
```


## Scatter Plot

The `ggplot` library works by letting users develop a plot layer by layer

- First, we associate a data to be plotted with the ggplot's canvas

```{r}
ggplot(surveys_complete)
```

- Next, we specify basic aesthetics of the plot. In this case, we include information about x- and 
y- axes. 

```{r}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length))
```


- To actually *draw* the plot, we specify the geometric of the plot

```{r}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point()
```


- The aesthetics of the geometric is separated from that of the data and ggplot canvas

```{r}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.4, aes(color = species_id))
```

- It is possible to customize the background layer (theme) and other information such as titles, axis labels, and subtitle. 

```{r}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.4, aes(color = species_id)) + 
  labs(subtitle="Weight versus Hindfoot Length", 
       y="Hindfoot Length (inches)", 
       x="Weight (grams)", 
       title="Scatterplot", 
       caption = "Source: Species Survey",
       color = "Species ID") + 
  theme_bw() +
  theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
        axis.text.y = element_text(colour="grey20", size=12),
        text=element_text(size=16, family="Arial"))
```

- Be careful with graphical fonts. 

```{r}
windowsFonts(Arial=windowsFont("Arial"))
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.4, aes(color = species_id)) + 
  labs(subtitle="Weight versus Hindfoot Length", 
       y="Hindfoot Length (inches)", 
       x="Weight (grams)", 
       title="Scatterplot", 
       caption = "Source: Species Survey",
       color = "Species ID") + 
  theme_bw() +
  theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
        axis.text.y = element_text(colour="grey20", size=12),
        text=element_text(size=16, family="Arial"))
```

- Make your graph portable across platforms:

```{r}
if (Sys.info()['sysname'] == 'Windows'){
  windowsFonts(Arial=windowsFont("Arial"))
}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.4, aes(color = species_id)) + 
  labs(subtitle="Weight versus Hindfoot Length", 
       y="Hindfoot Length (inches)", 
       x="Weight (grams)", 
       title="Scatterplot", 
       caption = "Source: Species Survey",
       color = "Species ID") + 
  theme_bw() +
  theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
        axis.text.y = element_text(colour="grey20", size=12),
        text=element_text(size=16, family="Arial"))
```


- Adjusting your labels: *hjust* and *vjust*:

```{r}
td <- expand.grid(
    hjust=c(0, 0.5, 1),
    vjust=c(0, 0.5, 1),
    angle=c(0, 45, 90),
    text="text"
)
str(td)
```

```{r}
ggplot(td, aes(x=hjust, y=vjust)) + 
    geom_point() +
    geom_text(aes(label=text, angle=angle, hjust=hjust, vjust=vjust)) + 
    facet_grid(~angle) +
    scale_x_continuous(breaks=c(0, 0.5, 1), expand=c(0, 0.2)) +
    scale_y_continuous(breaks=c(0, 0.5, 1), expand=c(0, 0.2))
```



- We can augment the plot with additional geometrics of various plots

```{r}
if (Sys.info()['sysname'] == 'Windows'){
  windowsFonts(Arial=windowsFont("Arial"))
}
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.4, aes(color = species_id)) + 
  geom_smooth(method = "loess", se = F) + # adding smoothing line
  labs(subtitle="Weight versus Hindfoot Length", 
       y="Hindfoot Length (inches)", 
       x="Weight (grams)", 
       title="Scatterplot", 
       caption = "Source: Species Survey",
       color = "Species ID") + 
  theme_bw() +
  theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
        axis.text.y = element_text(colour="grey20", size=12),
        text=element_text(size=16, family="Arial"))
```



```{r}
library(ggalt)
if (Sys.info()['sysname'] == 'Windows'){
  windowsFonts(Arial=windowsFont("Arial"))
}

medium_weight_length <- surveys_complete %>%
  filter(weight >= 15 & weight < 80 & hindfoot_length < 42 & hindfoot_length > 30)

ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.4, aes(color = species_id)) + 
  geom_smooth(method = "loess", se = F) + 
  geom_encircle(data = medium_weight_length,
                aes(x = weight, y = hindfoot_length), 
                color="red", 
                size=1, 
                expand=0.01) +   # encircle
  labs(subtitle="Weight versus Hindfoot Length", 
       y="Hindfoot Length (inches)", 
       x="Weight (grams)", 
       title="Scatterplot", 
       caption = "Source: Species Survey",
       color = "Species ID") + 
  theme_bw() +
  theme(axis.text.x = element_text(colour="grey20", size=12, angle=90, hjust=.5, vjust=.5),
        axis.text.y = element_text(colour="grey20", size=12),
        text=element_text(size=16, family="Arial"))
```


## Jitter Plot

- Is scatterplot adequate in this case?

```{r}
if (Sys.info()['sysname'] == 'Windows'){
  windowsFonts(Arial=windowsFont("Arial"))
}
ggplot(surveys_complete, aes(x = genus, y = hindfoot_length)) +
  geom_point()
```

- What can jitterplot do?

```{r}
if (Sys.info()['sysname'] == 'Windows'){
  windowsFonts(Arial=windowsFont("Arial"))
}
ggplot(surveys_complete, aes(x = genus, y = hindfoot_length)) +
  geom_jitter()
```

**Challenge**

Agument the jitter plot below with the following aesthetics to improve visibility:
- Color points by species
- Add legend for colors
- Turn x-axis label to 45 degree
- Change plot font to `Times New Roman`

```{r}
if (Sys.info()['sysname'] == 'Windows'){
  windowsFonts(Arial=windowsFont("____"))
}
ggplot(surveys_complete, aes(x = genus, y = hindfoot_length)) +
  geom_jitter()
```


## Correlogram

Quickly examine the correlation of multiple variables in the same dataframe. 

```{r}
library(ggcorrplot)
library(fueleconomy)

str(vehicles)
vehicles_numeric <- data.frame(cyl = as.numeric(vehicles$cyl),
                               displ = vehicles$displ,
                               hwy = as.numeric(vehicles$hwy),
                               cty = as.numeric(vehicles$cty))

vehicles_corr <- cor(vehicles_numeric, use = 'complete.obs')

ggcorrplot(vehicles_corr, hc.order = TRUE,
           type = 'lower',
           lab = TRUE,
           lab_size = 3,
           method = 'circle',
           colors = c("tomato2", "white", "springgreen3"),
           title='Correlogram of vehicles data',
           ggtheme = theme_bw)
```



